Tight Bounds for Dynamic Balanced Graph Partitioning

### 2020-05-09 paper/disc20clustering.tex
Line breaks

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\noindent
\textbf{Learning model.}
We study a variant of Online Balanced Partitioning,
where the communication pattern is \emph{static}:
an adversary determines the pairs of nodes that ever communicate, a priori,
and requests to these pairs arrive indefinitely.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al. \cite{sigmetrics19_partitioning}, we assume that the communication graph admits a \emph{perfect partition},
i.e., a partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

@@

A~\emph{ground set} is a group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
We say that a ground set is \emph{trivial}, if it contains exactly one node.
We say that the node that belongs to trivial a ground set is \emph{isolated}.

@@

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ isolated nodes exists.
	Fix the configuration just before the creation of a ground set with $x_i$ for some $i$.
	We show that if at least $\ell$ isolated nodes exist (including $x_i$), then there exists a perfect partition of ground sets after expanding a ground set with $x_i$.
	To show this claim, we consider a configuration, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this configuration, each cluster contains:  at most one ground set larger than $1$, possibly some isolated nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an isolated node, then we simply obtain a~perfect partition by swapping it with $x_i$.
	In the following we assume that no isolated nodes exist on $I(x_i)$.
	The largest ground set of $I(x_i)$ is at most $k-1$, because $x_i$ is on $S$.
	Therefore $\{x_0, y_0\}$ is present on $I(x_i)$, and we must evict it to make space for $x_i$.
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with isolated nodes, and we have at least $\ell$ isolated nodes --- hence a cluster with two isolated nodes exists.
	To find a feasible partition, we swap $\{x_0, y_0\}$ with these two collocated isolated nodes, and then we swap an isolated node from $I(x_i)$ with $x_i$.

@@

	\ALG performs at least one swap for every expansion of a ground set.
	Each expansion reduces the number of isolated nodes by $1$.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - l) / 2$.
\end{proof}

@@

In our algorithms, a~\emph{component} is a subset of frequently communicating nodes.
We maintain all nodes of each component collocated in one cluster~\cite{repartition-disc}.
If we need to move a node, we move the whole component that contains it.
We maintain a balanced partition of our components, a similar concept to partition of integers into sets with equal sum \cite{integer-partitions-book}.
In contrast, our partition is time-varying: communicating components are merged into one, and we adjust the partition.

@@

	Now we bound $\TAlg(\sigma_I)$.
	A~cluster of type $C_3$ contributes at most $3 \alpha - 1$ to $\TAlg(\sigma_I)$, as two pairs of nodes from the component are saturated and contribute $\alpha$ each, and the third, unsaturated pair contributes at most $\alpha-1$.
	Other cluster types contribute less: $C_1$ contributes $0$ and $C_2$ contributes $\alpha$.
	Summing this over all $\ell$ clusters gives us $\TAlg(\sigma_I) \leq (3 \alpha-1)\cdot \ell \leq 3\alpha\cdot\ell$.

@@

\subsection{Improved algorithm for Online Rematching} \label{sec:k2}
In this section,
we present \RM,
 an algorithm for the \OBRP restricted to clusters of capacity 2,
i.e.~$k=2$.
We interpret a pair of nodes collocated in one cluster as a matched pair.
Hence,
the problem is an online variant of the classic matching problem where
a matched pair can separate from each other in order to ``rematch'' with two other nodes.
This is known as the  \emph{Online Rematching} 
problem and a (non-strict) 7-competitive algorithm is already given by \cite{repartition-disc},
in which the ratio comes with an additive factor in $O(\alpha\ell^2)$.
We do not only improve upon their competitive ratio,
but also show that our ratio holds \emph{strictly}
(i.e., with no additive factor).
In addition,
our algorithm is similar
and slightly simpler than the one in \cite{repartition-disc}, 
while our analysis is significantly more simple and concise,
thanks to the charging scheme we devise here.

>>>>

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\noindent
\textbf{Learning model.}
We study a variant of Online Balanced Partitioning,
where the communication pattern is \emph{static}:
an adversary determines the pairs of nodes that ever communicate, a~priori,
and requests to these pairs arrive indefinitely.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al. \cite{sigmetrics19_partitioning}, we assume that the communication graph admits a \emph{perfect partition},
i.e., a partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

@@

A~\emph{ground set} is a group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a~partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
We say that a ground set is \emph{trivial}, if it contains exactly one node.
We say that the node that belongs to trivial a ground set is \emph{isolated}.

@@

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ isolated nodes exists.
	Fix the configuration just before the creation of a ground set with $x_i$ for some $i$.
	We show that if at least $\ell$ isolated nodes exist (including $x_i$), then there exists a~perfect partition of ground sets after expanding a ground set with $x_i$.
	To show this claim, we consider a configuration, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this configuration, each cluster contains:  at most one ground set larger than $1$, possibly some isolated nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an isolated node, then we simply obtain a~perfect partition by swapping it with $x_i$.
	In the following we assume that no isolated nodes exist on $I(x_i)$.
	The largest ground set of $I(x_i)$ is at most $k-1$, because $x_i$ is on $S$.
	Therefore $\{x_0, y_0\}$ is present on $I(x_i)$, and we must evict it to make space for $x_i$.
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with isolated nodes, and we have at least $\ell$ isolated nodes --- hence a~cluster with two isolated nodes exists.
	To find a feasible partition, we swap $\{x_0, y_0\}$ with these two collocated isolated nodes, and then we swap an isolated node from $I(x_i)$ with $x_i$.

@@

	\ALG performs at least one swap for every expansion of a ground set.
	Each expansion reduces the number of isolated nodes by $1$.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - l) / 2$.
\end{proof}

@@

In our algorithms, a~\emph{component} is a subset of frequently communicating nodes.
We maintain all nodes of each component collocated in one cluster~\cite{repartition-disc}.
If we need to move a~node, we move the whole component that contains it.
We maintain a balanced partition of our components, a similar concept to partition of integers into sets with equal sum \cite{integer-partitions-book}.
In contrast, our partition is time-varying: communicating components are merged into one, and we adjust the partition.

@@

	Now we bound $\TAlg(\sigma_I)$.
	A~cluster of type $C_3$ contributes at most $3 \alpha - 1$ to $\TAlg(\sigma_I)$, as two pairs of nodes from the component are saturated and contribute $\alpha$ each, and the third, unsaturated pair contributes at most $\alpha-1$.
	Other cluster types contribute less: $C_1$ contributes~$0$ and $C_2$ contributes $\alpha$.
	Summing this over all $\ell$ clusters gives us $\TAlg(\sigma_I) \leq (3 \alpha-1)\cdot \ell \leq 3\alpha\cdot\ell$.

@@

\subsection{Improved algorithm for Online Rematching} \label{sec:k2}
In this section,
we present \RM,
 an algorithm for the \OBRP restricted to clusters of capacity $k=2$.
We interpret a pair of nodes collocated in one cluster as a matched pair.
Hence,
the problem is an online variant of the classic matching problem where
a matched pair can separate from each other in order to ``rematch'' with two other nodes.
This is known as the  \emph{Online Rematching} 
problem and a (non-strict) 7-competitive algorithm is already given by \cite{repartition-disc},
in which the ratio comes with an additive factor in $O(\alpha\ell^2)$.
We do not only improve upon their competitive ratio,
but also show that our ratio holds \emph{strictly}
(i.e., with no additive factor).
In addition,
our algorithm is similar
and slightly simpler than the one in \cite{repartition-disc}, 
while our analysis is significantly more simple and concise,
thanks to the charging scheme we devise here.
